Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-28 resubmission/04-analisis.qmd
update discussion and conclusion (no render)

```{r echo=FALSE, results='asis'}
#| label: tbl-interactions2
#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
#| tbl-cap-location: top

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time has no statistically significant effects on either the within- or between-person effects of perceived inequality and meritocratic perceptions. The within-person interaction effects are positive, suggesting that in later waves, the influence of perceived inequality and meritocratic beliefs on market justice preferences tends to increase. Conversely, in the between-person specification, the interaction effects are negative, indicating that over time, the impact of perceived inequality and meritocratic beliefs on support for market-based justice tends to diminish. However, these trends are not statistically significant, and thus no definitive conclusions can be drawn.

>>>>

```{r echo=FALSE, results='asis', eval=FALSE, include=FALSE}
#| label: tbl-interactions2
#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
#| tbl-cap-location: top

@@

Regarding the temporal dynamics of the key predictors, the analysis shows that time has no statistically significant effects on the intrapersonal or interpersonal effects of perceived inequality and meritocratic perceptions. Further details about this analysis can be found in supplementary material.

### 2025-07-28 resubmission/05-discussion.qmd
Update carpeta resubmission y varios (archivos no renderizados)
update discussion and conclusion (no render)

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Interestingly, the negative within-effects observed over time implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. Although this moderation is not significant for within-subject effects (i.e. over time), when taking into account between-subjects differences we found a positive effect. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic perceptions, perceptions of inequality, and attitudes toward market justice. The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms. Post-mobilization, the positive relationship between meritocratic perceptions and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives. These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness. This is aligned with literature emphasizing that social protests act as catalysts for transforming societal attitudes by disrupting normative beliefs and encouraging more critical evaluations of institutional legitimacy [@tilly_contentious_2015]. The findings underscored the importance of socio-political context in shaping and reshaping attitudes toward inequality and justice, highlighting how collective action could temporarily or permanently alter the normative foundations that sustained support for market justice.

>>>>

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed less support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also further foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021]. Interestingly, intraindividual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects. While an increase in the perception that rewards are based on individual effort is associated with stronger preferences for market fairness, an increase in the belief that rewards come from talent is linked to lower support for such principles. One possible explanation is that effort is generally perceived as a controllable and malleable trait, which aligns with the normative foundations of market fairness. Conversely, talent may be viewed as an innate and less alterable attribute, making talent-based reward allocation seem less fair or legitimate. Moreover, greater exposure to real-life situations where outcomes are strongly determined by inherent characteristics, rather than effort or hard work, may lead people to question the fairness of market-based reward systems.

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to strengthen when perceptions of economic inequality were particularly high. While this moderation is not significant for effort-related effects, it is significant for talent in the within- and between-individual effects. This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases. Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative. Therefore, as perceived inequality become less negative (i.e. more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences. This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.

Finally, the fourth hypothesis addressed the impact of recent social movements—particularly those that emerged in 2019—on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice. The main effect of time suggests that, on average, support for market justice increased after 2019. However, the results do not reveal any significant interactions between distributive mechanisms and market justice attitudes, indicating that the 2019 social mobilizations did not substantially alter the underlying associations between these variables. This may suggest a certain stability or resilience in the normative frameworks that shape individuals’ evaluations of justice, even in the context of major collective political events. Alternatively, it is possible that while the protests challenged institutional legitimacy, they did not fundamentally shift individuals’ views regarding how rewards should be allocated in society.

### 2025-07-28 resubmission/06-conclusion.qmd
Update carpeta resubmission y varios (archivos no renderizados)
update discussion and conclusion (no render)

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic perceptions strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic perceptions, such that as the awareness of disparities are attenuated (i.e. less negative) the role of meritocratic perceptions on existing market-based allocations become stronger. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

>>>>

The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. At the same time, meritocratic beliefs—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that these beliefs serve as moral justifications for structural inequalities. However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it. This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits. Moreover, the interaction between inequality perceptions and meritocratic beliefs indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities. This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.
